---
title: "Global Invasive and Alien Traits and Records (GIATAR) Database - Query Functions Tutorial"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

# Global Invasive and Alien Traits and Records (GIATAR) Database - Query Functions Tutorial

Welcome to the tutorial for the query functions supplied with the GIATAR database! The file containing query functions `GIATAR_query_functions.R` is available in the queries folder of the supporting GitHub repository and in the queries folder of the released database folder (This tutorial should be stored adjacently).

These functions simplify the process of querying and joining the database and typically return data frames, to simplify the process of analysis. While there is considerably more information in the database than is accessible through these tools, we hope they will simplify the most common operations for database users.

## Environment

The environment setup file supplied with the code for this project will suffice for the query functions here. However, it contains a complete suite of packages for database updating, some of which may be tricky to install. These query functions rely mostly on basic R packages (`dplyr`, `tidyr` etc.) with the exception of `rgbif`, which can be installed from CRAN.

## Setting Paths

### To functions

We suggest putting the `GIATAR_query_functions.R` file (available in the queries folder of the DOI released database or on the project GitHub `https://github.com/ncsu-landscape-dynamics/GIATAR-database` for the most current version) in your project directory.

### To data

`GIATAR_query_functions.R` contains `data_path` as the first line of code in the file - set this to the database directory of GIATAR. If you prefer, you can call `create_dotenv(pathtodata)` to create a `.env` file that permanently sets this path.

## usageKeys

Unique ID Keys for species in the database are referred to as `usageKeys`, following the structure and naming of usageKeys from GBIF. Where possible, we have retained GBIF usageKeys as unique IDs for taxa - otherwise, we have generated unique IDs that won't overlap with GBIF usageKeys.

## Basic Operations

```{r setup, include=FALSE}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(rgbif)
source("query_functions/R/GIATAR_query_functions.R")

## Set the path to data

```


```{r run-tutorial, include=FALSE}
# Basic operations
# Get all species
get_all_species()


# Check if species exists and get usageKey
check_species_exists("Ailanthus altissima")
get_usageKey("Ailanthus altissima")

# Pull some introduction records
get_first_introductions("Apis mellifera", import_additional_native_info = TRUE)

# Pull ecology information
get_ecology("Thrips tabaci")

# Get hosts and vectors
get_hosts_and_vectors('Icerya purchasi')

# Get native ranges
get_native_ranges("Apis mellifera")

# Get common names
cnms <- get_common_names("Pancratium maritimum")

# Get usageKey
get_usageKey("Pancratium maritimum")

# Get trait table list
get_trait_table_list()

# Get trait table
get_trait_table('CABI_airtemp')
```

