---
title: "Global Invasive and Alien Traits and Records (GIATAR) Dataset - Query Functions Tutorial"
author: "Thom Worm and Ariel Saffer"
date: "2024-07-22"
output: html_document
---

# Global Invasive and Alien Traits and Records (GIATAR) Dataset - Query Functions Tutorial

Welcome to the tutorial for the query functions supplied with the GIATAR dataset! The file containing query functions `GIATAR_query_functions.R` is available in the queries folder of the supporting GitHub repository and in the queries folder of the released dataset folder (This tutorial should be stored adjacently).

These functions simplify the process of querying and joining the dataset and typically return data frames, to simplify the process of analysis. While there is considerably more information in the dataset than is accessible through these tools, we hope they will simplify the most common operations for dataset users.

## Environment

The environment setup file supplied with the code for this project will suffice for the query functions here. However, it contains a complete suite of packages for dataset updating, some of which may be tricky to install. These query functions rely mostly on basic R packages (`dplyr`, `tidyr` etc.) with the exception of `rgbif`, which can be installed from CRAN.

## Setting Paths

### To data

The .env file contains a variable called `DATA_PATH` - set this to the dataset directory of GIATAR. If you prefer, you can call `create_dotenv(pathtodata)` to create a `.env` file that permanently sets this path.

## usageKeys

Unique ID Keys for species in the dataset are referred to as `usageKeys`, following the structure and naming of usageKeys from GBIF. Where possible, we have retained GBIF usageKeys as unique IDs for taxa - otherwise, we have generated unique IDs that won't overlap with GBIF usageKeys.

## functions
returns species name as string - takes usageKey as string or int

### get_GIATAR_current(data_dir)
returns the current version of the GIATAR dataset to the directory provided.

This function fetches the current version of the GIATAR dataset, ensuring that users have access to the most recent data available at the time of the query. 

### get_usageKey(species_name) 
 returns usageKey as string - takes species name as string

### get_all_species() 
 returns list of all species names in dataset - no inputs

### check_species_exists(species_name) 
 returns True or False - takes species name as string

### get_first_introductions(usageKey, check_exists=False, ISO3_only=False, import_additional_native_info=True) 


returns dataframe of first introductions takes usageKey as string or int

check_exists=True will raise a KeyError if species is not in dataset

ISO3_only=True will return only return species location info that are 3 character ISO3 codes. Some other location info includes bioregions or other geonyms

import_additional_native_info=True will import additional native range info, first by seeing if native range info for a particular country is availible from sources that reported later than the first introduction, and second by importing native range info from the file of native range info unique to GIATAR

### get_all_introductions(usageKey, check_exists=False, ISO3_only=True) 

returns dataframe of all introductions - takes usageKey as string or int 
check_exists=True will raise a KeyError if species is not in dataset
ISO3_only=True will return only return species location info that are 3 character ISO3 codes. Some other location info includes bioregions or other geonyms

import_additional_native_info=True will import additional native range info, first by seeing if native range info for a particular country is availible from sources that reported later than the first introduction, and second by importing native range info from the file of native range info unique to GIATAR

### get_ecology(species_name) 
 returns dictionary of dataframes of ecology info - takes species name as string. Ecology info returned by this function includes rainfall, airtemp, climate, lat/altitude, water temp and wether a pest utilizing wood packaging.

 Ecological info is variously formatted for different species - e.g. air temperature might include max, min, range or other info. We reccomend spending time with the outputs to find the information you want. 



### get_hosts_and_vectors(species_name) 
returns dictionary of dataframes of host and vector info - takes species name as string
This tool returns hosts (plant hosts for herbivorous insects, animal hosts for diseases and parasites) and vectors (either zoonotic or plant vectors - mostly for diseases)



### get_species_list(kingdom=None, phylum=None, taxonomic_class=None, order=None, family=None, genus=None)
 returns list of usageKeys matching taxonomic criteria - takes kingdom, phylum, taxonomic_class, order, family, genus as strings. This function can help select a group of organisms in the dataset matching the search criterion. Note that the term <code>class</code> is protected in python, so we refer to the taxonomic grouping as taxonomic_class



### get_native_ranges(usageKey, ISO3=None) 
returns dataframe of native ranges - takes usageKey as string or int.

The GIATAR dataset stores native range information in several ways - some better-studied species have native information as a binary true/false for the country level. Many other species have native range information stored only as biogeographic zones e.g. palearctic. We provide functionality to map this biogeographic zone data to presence-absence t/f using a crosswalk, which is availible in the native ranges subfolder of the dataset.

When the user calls <code>get_native_ranges()</code> and ISO3 is set to None, the function returns all avalible information about the species. 
If the user wishes to use the native-range to country-presence crosswalk, they should provide ISO3 as a python list of ISO3 standard country codes e.g. <code>['USA','CHN]</code> - the function will then use the crosswalk to provide true/false information on the native status of the species if there is biogeographic information avalible. When biogeographic information and more specific country/native binary information is avalible, the function defaults to the more specific country true/false info. 

ISO3=list returns dataframe of native ranges and True or False if species is native to ISO3 - takes a list of ISO3 codes for countries as input. See examples below for context.

## Functions to seek a list of invasive species

### get_taxa_by_host(host_name)
returns a list of usageKeys for taxa that use the specified host - takes the host name as a string. This function helps identify all taxa in the dataset that are associated with a particular host species. 

For example, if you want to find all species that use "Citrus" as a host, you can use this function to get a list of their usageKeys. This is particularly useful for researchers looking to study the suite of invasive species affecting or traveling with a given host species.

### get_taxa_by_pathway(pathway_name)
returns a list of usageKeys for taxa that are associated with the specified pathway - takes the pathway name as a string. This function helps identify all taxa in the dataset that are associated with a particular pathway.

For example, if you want to find all species that are introduced through "Horticulture" as a pathway, you can use this function to get a list of their usageKeys. This is particularly useful for researchers looking to study the suite of invasive species introduced through a specific pathway.

### get_taxa_by_vector(vector_name)
returns a list of usageKeys for taxa that are associated with the specified vector - takes the vector name as a string. This function helps identify all taxa in the dataset that are associated with a particular vector.

For example, if you want to find all species that are transmitted by "Aedes aegypti" as a vector, you can use this function to get a list of their usageKeys. This is particularly useful for researchers looking to study the suite of invasive species transmitted by a specific vector.

## Basic Operations

```{r setup, include=FALSE}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(rgbif)
source("query_functions/R/GIATAR_query_functions.R")

## Set the path to data
```

```{r tutorial, include=FALSE}
# Basic operations
# Get all species
get_all_species()


# Check if species exists and get usageKey
check_species_exists("Ailanthus altissima")
get_usageKey("Ailanthus altissima")

# Pull some introduction records
get_first_introductions("Apis mellifera", import_additional_native_info = T)

# Pull ecology information
get_ecology("Thrips tabaci")

# Get hosts and vectors
get_hosts_and_vectors("Icerya purchasi")

# Get native ranges
get_native_ranges("Apis mellifera")

# Get common names
cnms <- get_common_names("Pancratium maritimum")

# Get usageKey
get_usageKey("Pancratium maritimum")

# Get trait table list
get_trait_table_list()

# Get trait table
get_trait_table("CABI_airtemp")
```

## Querying Invasive Taxa by Traits

You can also query invasive taxa by traits. This lets you then get records for a suite of invaders. Here are some examples.

```{r invasive_taxa_by_traits}
# Example 1: Querying by pathway
pathway_name <- "ornamental"
taxa_by_pathway <- get_taxa_by_pathway(pathway_name)
print(taxa_by_pathway)

# Example 2: Querying by vector
vector_name <- "Aedes aegypti"
taxa_by_vector <- get_taxa_by_vector(vector_name)
print(taxa_by_vector)

# Example 3: Querying by host
host_name <- "mays"
taxa_by_host <- get_taxa_by_host(host_name)
print(taxa_by_host)
```

```{r introductions_by_trait}
# Create a table of first introductions for all taxa in the taxa_by_host example
first_introductions_df <- data.frame()
for (taxon in taxa_by_host$canonicalName) {
    # print(taxon)
    df <- get_first_introductions(taxon)
    df$canonicalName <- taxon
    first_introductions_df <- bind_rows(first_introductions_df, df)
}
```